---
title: "Class_6"
format: html
---

# Class
## Dataimport
```{r}
library("readr")
library("sf")


wildschwein_BE <- read_delim("Data/wildschwein_BE_2056.csv", ",") |>
    st_as_sf(coords = c("E", "N"), crs = 2056, remove = FALSE)
```

IMporting wildboar data from csv 
<<- Very generic data format
- reasonable for points... 

But to import a geopackage, we can use st_read/ read_sf
```{r}
feldaufnahme <- st_read("Data/Feldaufnahmen_Fanel.gpkg")
```

The import gives us some metadata, which we should give a wuick look

## Dataexploration
```{r}
View(feldaufnahme)

# The condole gives us more information 
feldaufnahme
str(feldaufnahme)

plot(feldaufnahme)
# creates on plot per column (weird)
# to Just plot one column 
plot(feldaufnahme["Frucht"])
```

The plot function is much faster in computation time than ggplot. It is thus very usefull for exploration... 

## Join datasets
### VIsually
```{r}
library(ggplot2)
library(lubridate)
library(dplyr)

# filter wildboardata to only use data for which there is existing crop data
wildschwein_BE
mayjune <- wildschwein_BE |> 
  mutate(month = month(DatetimeUTC)) |> 
  filter(month >= 5, month <= 6)

ggplot() +
  geom_sf(data = feldaufnahme,
          aes(fill= Frucht)) +
  geom_sf(data= mayjune,
          aes(shape = TierName), alpha=.1 )+
  theme(legend.position = "none")
```

## Annotate Wilboar data with crop information
```{r}
?st_join

joined <- st_join(mayjune, feldaufnahme) #this join doesn't need by=, it uses the geometry automatically
View(joined)
```

When Wildboar is outside of feldaufnahme area, the FIeld ID and Frucht is NA
When Wildboar is not inside an agricultural field, the FieldID = 0 and the Frucht= Wald/Feuchtgebiet

## VIsulize
```{r}
library(forcats) #very good library for dealing with factor levels (used for lumping)

joined |> 
  mutate(
    hour = hour(DatetimeUTC),
    Frucht = fct_lump(Frucht, 5)) |> #lumps the Frucht category to the 5 most dominant (do this make plot more readable )
  st_drop_geometry() |> 
  count(TierName, hour, Frucht) |> 
  # How many times was rosa in an Acker
  group_by(TierName, hour) |> 
  mutate(
    n_total = sum(n), 
    perc = n/n_total) |> 
  #make the plot
  ggplot(aes(hour, perc, color= Frucht)) +
  geom_line() +
  facet_wrap(~TierName)
 
```

## Use a different join method than intersect

You can look up different join functions in help!
```{r}
joined2 <- st_join(mayjune, feldaufnahme, join = st_within) #this join doesn't need by=, it uses the geometry automatically
View(joined2)

joined2 |> 
  mutate(
    hour = hour(DatetimeUTC),
    Frucht = fct_lump(Frucht, 5)) |> #lumps the Frucht category to the 5 most dominant (do this make plot more readable )
  st_drop_geometry() |> 
  count(TierName, hour, Frucht) |> 
  # How many times was rosa in an Acker
  group_by(TierName, hour) |> 
  mutate(
    n_total = sum(n), 
    perc = n/n_total) |> 
  #make the plot
  ggplot(aes(hour, perc, color= Frucht)) +
  geom_line() +
  facet_wrap(~TierName)

```

## Rasterdata
```{r}
# load raster data
library(terra)


veg <- rast("Data/vegetationshoehe_LFI.tif")
# veg <- terra:: project(veg, "epsg:2056") 
# this line of code ensures, that the raster data is in the same projection as our wildboar data

plot(veg)

#extract information from the vegetationshähe to your wildboar data
?extract

extract <- extract(veg, joined) #the warning 
joined$vegetationshoehe <- extract$vegetationshoehe_LFI
```

### Visualizing added raster data
```{r}

joined |> 
  mutate(
    hour = hour(DatetimeUTC)) |> 
  group_by(TierName, hour) |> 
  summarise(
    mean = mean(vegetationshoehe, na.rm = T)
  ) |> 
  ggplot(aes(hour, mean, color = TierName)) +
  geom_line() 

```

